
\section{Outlook}
\label{sec:outlook}

This review shows various possibilities for future work within the specific contexts of creative pattern generation. Next to directions discussed in ~\Cref{sec:discussion_creative_means}, we would like to highlight directions that we haven't mentioned yet but which have great potential in regard to creative control mechanisms.

Collaboration is a valuable future line of investigation for enabling creative work. More aspects of common tools are either fully browser-based, or are in some way connected to the cloud-based storage of assets, settings and results; therefore, they function online and are easily shared. Collaboration is closely connected to the previously discussed issue of navigation histories. This is not only relevant for individual work processes but also for more general production pipelines in a commercial context. In this regard, the sharing and collaborative work on iterations, which involves multiple persons referencing different versions, is essential. Work has been done (\eg~\cite{talton_2009_emw, salvati_2015_mcm,oleary_2018_csi}) but further investigations of collaboration for creative control are called for. 

The procedural content generation (PCG) community for games is pushing the general integration of artificial intelligence (AI) into a procedural creation process. A new discipline of \textit{mixed-initiative creative interfaces} is rising as, \eg~an ACM CHI workshop under the same name in 2017 shows~\cite{deterding_2017_mci}. The workshop summary states, it is the goal to ``put human and computer in a tight interactive loop where each suggests, produces, evaluates, modifies, and selects creative outputs in response to the other.'' In order to achieve this, AI enables computer agency, and novel interfaces enable collaboration between computers and human users. In regard such AI systems and the integration of agency and automation also \citeauthor*{heer_2019_apa}~\cite{heer_2019_apa} discusses three case studies and poses several open questions for further developments.

The usage of semantic attributes presents a highly intuitive navigation technique, which so far has been successfully applied in the context of shape modifications, for example by \citeauthor*{yumer_2015_sse}~\cite{yumer_2015_sse}. For the control of procedural textures, methods are based on the analysis and description of texture in regard to human perception, which has a long research tradition, starting with textons from \citeauthor*{julesz_1981_tte}~\cite{julesz_1981_tte}. 
Since then, this line of research has continued, and texture descriptions with perceptual \cite{liu_2015_vpp} and semantic~\cite{matthews_2013_eta,cimpoi_2014_dtw} attributes have been investigated. \citeauthor*{dong_2017_ptg}~\cite{dong_2017_ptg} and \citeauthor*{liu_2018_ppt}~\cite{liu_2018_ppt} employed such features for the navigation of a procedural texture space and for the generation of suitable textures by given features. However, the results of such studies are still limited and of varying quality - and the authors themselves \cite{liu_2018_ppt} call their results experimental. Nonetheless, these works present an interesting approach that is worth further investigation, especially in combination with machine learning algorithms. Because many pattern designs are structured and follow an internal logic, it seems feasible to come up with a collection of suitable attributes. For example an ornamental design space is much smaller in comparison to all ``textures in the wild''~\cite{cimpoi_2014_dtw}, and ornamentation could constitute a valuable context for further investigations into the incorporation of semantic attributes into a creative creation process.

As discussed in \Cref{sec:models}, control techniques are closely inter-related with the representation of the underlying models. 
% Therefore, a more unified development of models across research communities would be beneficial. Expert knowledge of usability should especially be considered. However, 
Further automation for the creation of complex models also poses interesting challenges, such as abstraction \cite{nan_2011_cgr}, symmetry computation~\cite{cullen_2011_sh} and design space variations.