%!TEX root = ../deco_star.tex


\section{Creative Means}\label{sec:creativ_means}

% ZOOM INTO TOPIC
Many content generation contexts in Computer Graphics involve creative considerations. To consider creativity is especially relevant for investigating control mechanisms because the controls provide the means for creative creation. However, creativity is also a notoriously difficult topic to address, because it is ill-defined and involves insights from various disciplines.

On the one hand, there are efforts to develop algorithms that perform creatively for creativity-related research in computer science. On the other hand, there is the common goal of supporting human creativity with digital tools, which is the focus of this survey. The goal of enabling human creativity is, for example, well-researched in regard to interface design. This is an established field in the human-computer interaction (HCI) community and it is based largely on the pioneering work of \citeauthor*{shneiderman_2007_cst}~\cite{shneiderman_2007_cst} about \textit{Creativity Support Tools}. This topic has recently found attention again in the HCI community~\cite{frich_2018_tyo, frich_2019_mtl, remy_2020_ecs}, asking for more rigorous evaluation in regard to claimed creativity support. Among other aspects,~\citeauthor*{frich_2018_tyo}~\cite{frich_2018_tyo} ask for "clearer definitions of creativity" and report that in 113 surveyed publications, "only 26 papers directly quoting creativity metrics (\eg fluency, originality, novelty, flexibility)." This shows that the HCI community already strives for the type of survey this STAR proposes. However, for this survey, rooted in the field of computer graphics, we review potentially relevant characteristics of underlying algorithms that might help to enable artists to be creative. Interface design aspects are included but are not focus of this survey. However, the need for bridging between developing the core algorithms and the interface in order to support creative-artistic intent has been voiced by past research~\cite{deterding_2017_mci,isenberg_2016_inw, salesin_2002_nar}, and this work wishes to contribute to this effort.

\citeauthor*{cherry_2014_qcs}~\cite{cherry_2014_qcs} presented an evaluation technique called the quantifiable \textit{Creativity Support Index} (CSI), which has found its way into the graphics community~\cite{shugrina_2017_ppi}. The index measures how well a tool enables creativity based on a psychometric survey. The development and validation of the measurement dimensions -namely, \textit{exploration}, \textit{expressiveness}, \textit{immersion}, \textit{enjoyment}, \textit{results worth effort}, and \textit{collaboration}- are based on user tests. \citeauthor*{cherry_2014_qcs}~\cite{cherry_2014_qcs} quantified the specific phrases participants used to describe a creative process. However, a clear definition of terms like \textit{exploration} and \textit{expressiveness} is missing or the meaning of a statement such as ``I was able to be very creative, [...]'' is left open. 

A quantified user study is often not feasible, for example for this survey and for assessing the state of the art. Furthermore, doing so would not be meaningful because the support of creativity is not a goal for most methods. However, most methods do offer carefully developed control mechanisms. We propose it to build a discussion of the means for creativity on the presented control mechanisms in a publication and on the specifics given from the authors. Based on the given information, we reflect on the potential for creative means in a meaningful way, even if creative control was not necessarily the authors' intention. 

We derive a working basis of such creative means in regard to control mechanisms in the following review of relevant references. Our classification is meant as a step toward understanding the creative control options within the current state of the art. In terms of measurement dimensions, it can be seen as an evaluation on algorithmic level and a subset of the more general and user-study-based classification with the \textit{Creativity Support Index}. 

\subsection{Discussion Basis}\label{subsec:creative_means}
% Academic discussions about what constitutes creativity have a long history in the field of psychology~\cite{weisberg_2006_cui}, cognitive science~\cite{boden_2004_cmm} and philosophy~\cite{gaut_2010_pc} and is ongoing. 

% TODO: Check tense - should it be in present?
%BB - I believe present is OK. 

\citeauthor*{weisberg_2006_cui}~\cite{weisberg_2006_cui} states that a creative person ``intentionally produces a novel product'' (p.70) and explicitly decouples a possible value of a product from it being the result of a creative process (p.63). 
% We follow Weisberg's exclusion of a often diffuse value of the result of a creative process and focus on the creative intent. 
\citeauthor*{boden_2010_cat}~\cite{boden_2010_cat} described \textit{novelty} as a surprising product, one that the creator did not directly anticipated (p.30). 
% \citeauthor*{boden_2010_cat} also differentiated between a product being surprising or novel to oneself in contrast to something being universally novel (\cite{boden_2010_cat} p.30). In this survey of creativity regarding control mechanisms, we only include novelty in reference to the expectations of a single artist.

The integration of \textit{intention} in describing a creative process is crucial for the development of meaningful algorithms. Weisberg explains that a painter who accidentally stains a painting - a stain which is later applauded by the art world as an innovative technique - cannot be considered to have achieved a creative result. Hence, algorithms need to enable artists to follow their intentions with transparent and controllable mechanisms. 
% The idea, for example, of an algorithm producing a large number of random design choices for an artist to choose from contradicts this principle of intention.

Weisberg argues that a creator needs domain-specific knowledge and expertise in order to come up with something novel or surprising. He rejects the common perception of creativity as being an ``unfathomable leap of insight'' and advocates its systematic accessibility. 
% He argues that the perception of creativity as unfathomable results from missing context and domain knowledge and from neglecting to include the whole process that leads to a novel product. 
Weisberg's argument applied in the context of control mechanisms leads to requiring techniques to enable artists to fully understand the domain they work with. Cause and effect of interactions as well as the overall options to control the output must be transparent and navigable.
% Weisberg concludes his considerations about creative processes by stating that ``you must also work to broaden and deepen your database''~\cite{markman_2009_tis}. Hence, 
% control mechanisms not only need to be transparent and fully steerable for an artist, 

However, there also must be a large design space for a creator to explore. %\citeauthor*{boden_2010_cat}~\cite{boden_2010_cat} describes this as a landscape to navigate through. 
Such an increase of possible options is a core aspect of many common creativity techniques, such as brainstorming, and must also be used for the development of digital tools~\cite{terry_2004_vea}. Design spaces need to be big but also meaningful and well framed for the domain it represents. It has to provide space to delve into without the danger of getting lost. 
% Classical brainstorming, for example, is on the one hand based on the idea of coming up with as many answers to a question as possible - with no restrictions. On the other hand, a brainstorming session starts with a carefully crafted problem statement, which is supposed to be as precise and descriptive as possible. Hence, human brainstormers intuitively remain in the domain of the problem statement and exclusively offer solutions related to the problem. 
Therefore in a system that offers design options, all options need to make sense, while ``enabling someone to see possibilities they hadn't glimpsed before''~\cite{boden_2010_cat}. For the discovery of unseen results various stimuli can be given, for example well designed constraints can act as such stimuli. 

% Common creativity techniques often include stimulating constraints~\cite{onarheim_2010_occ,shih_2011_buc,biskjaer_2014_cud,stokes_2005_ccp} or motivations to guide the exploration in a specific direction. For example, with the Six Thinking Hats technique \cite{debono_1985_sth}, each hat represents a specific mindset, such as "critical" or "emotional", with which a participant should operate. This technique enables a large variety of possible stimuli and cues such as associations, analogies, abstractions, visualizations and reversals, including purely random inputs. Stimuli are a field of active research and as mentioned above, the usefulness of random cues is doubted. \citeauthor*{markman_2009_tis} summarize their insights (\cite{markman_2009_tis}, pp.48-69) about cognitive support for creative processes, saying that the pool of random stimuli needs to be restricted to increase the opportunity for novelty and to decrease the probability of misleading failures (p.68).

In a process to enable creativity, the target audience is also an influencing factor. Each skill level requires its own unique type of support~\cite{cherry_2014_qcs}. 
%\citeauthor*{cherry_2014_qcs}~\cite{cherry_2014_qcs} discuss the handling of different user competencies as example for the importance of balancing simplicity and expressiveness. The authors mention that more expressive but also more complex tools score higher on the CSI. 
In the following discussions, this survey does not explicitly investigates the appropriateness of a technique for different skill levels (unless it specifically distinguishes a related work, as, for example, in \cite{benedetti_2014_pba}) but instead focuses on the general suitability of a technique to create the design goal with a reasonable training curve for an average artist. Also, for this first investigation, we focus on support for a single artist but it is worth mentioning that an enabling of collaborative creative work would be a meaningful aspect to add in the future.
% collaboration as big next question
% http://ceur-ws.org/Vol-1907/7_mici_dove.pdf
% http://ceur-ws.org/Vol-1907/6_mici_dalsgaard.pdf

To sum up the brief review in the context of control mechanisms, we specify the categories \textit{navigation}, \textit{transparency}, \textit{variation} and \textit{stimulation} in the context of control mechanisms. We understand variation as the size of the design space within the context of the technique. For the exploration of different designs we distinguish between the general controllability necessary for navigating a design space (``there are many different roads in the landscape''), and the transparency of that navigation and the understanding of cause and effect when using the tool (``I have the map to the landscape and know how to get from one point to another'').  However, at this point, these categories can be seen as somewhat loose and experimental, aiming toward a better understanding of requirements for creative controls. For these characteristics, there is no clear translation into quantifiable metrics, such as timings or error rates, which are standardized measurements for productivity~\cite{cherry_2014_qcs,shneiderman_2007_cst}.

\noindent\textbf{Navigation}

The means of navigation describe whether a creation processes is efficiently manageable as well as the extent of the controllability.

\begin{itemize}
    \item \textit{Interactive:} Refers to a system with ideally no noticeable delays when executing controls and computing results. Lengthy, non-creative configuration requirements are also potentially distracting. Hence, a thorough analysis should consider the whole process an artist has to go through to produce a result.
    \item \textit{Quantity of Controls:} This category indicates how flexible and controllable a technique is, \eg~by counting the number of different controls that can be adjusted for one output. 

    Ideally, this category would refer to the ratio of visual features of the possible output that are relevant to humans to controllable features. This would ensure that the controls cover all necessary features and that they complement each other. However, the identification of generally describable, perceptually relevant visual features is out of the scope of this survey and left to future work.

    \item \textit{Navigation History:} Describes the ability to go back and forth in one's own creation process, such as using an eraser.
\end{itemize}



\noindent\textbf{Transparency}
\label{subsubsec:transparency}

The means of transparency describe how clear the understanding of cause and effect within the system are.

\begin{itemize}
    \item \textit{Control Domain:} Refers to how well controls are mapped to visual features and how well they cover the possible design range of each feature. A high-quality control should not have any overlapping effects with other controls.
    \item \textit{Control Communication:} This category describes how well controls (\eg~with a visualization and/or little abstraction) represent their effects on the result. For artist-centered tools this could mean that controls should be visual and directly on the canvas.
\end{itemize}


\noindent\textbf{Variation}\label{subsubsec:variation}

Variation indicates how visually different the results can be.

\begin{itemize}
    \item \textit{Size of the Design Space:} is limited if all results look rather similar to each other and are part of a specific design class. A large design space of one technique allows, for example, for different texture classes such as combining stochastic and structural creation.
    \item \textit{Openness of the Design Space:} Refers to the limitlessness of possible designs and that there is no attachment of the technique to a specific design class. An open design space enables an artist to come up with a distinctive individual style, for example. Different artists can create inherently different and unique results with the same tool if it has a open desgin space.
\end{itemize}

We do understand that a clear definition of the available different design classes is needed. However, these solely dependent on the design context. 


\noindent\textbf{Stimulation}
\label{subsubsec:stimulation}

The means of stimulation indicate how well an artist can enter a pleasurable and stimulating workflow.
\begin{itemize}
    \item \textit{Immersion:} How natural and enjoyable the usage of a system feels.

    An immersive technique needs to be fluent to navigate, controls have to be intuitive and the design space large enough to not to hit its boundaries while using the tool. 
    
    \item \textit{Stimuli:} The support to find surprising results - for example, with design suggestions or variations of the input. Options to support stimulation are still underrepresented but on the rise with machine learning techniques. A clear definition of this category is not feasible at this point.

\end{itemize}

% For stimulation being more broadly accepted as relevant research question and required element for tools that enable creative creation, first more research has to go into understanding what constitutes effective stimulation before it can be implemented. A psychologically well-grounded theoretical assessment of stimuli is out of scope of this thesis. For the following discussion stimuli are included if a technique is explicitly evaluated in regard to its stimuli, for example with a survey. However, this category can not be included overall and in a unified manner and its generalizable assessments is left to future work.

% \subsection{Summary}
% \label{subsec:creativity_summary}

In summary, for handling the ill-defined topic of creativity, we follow the definition of creativity as intentionally producing a novel and surprising product. We derive means for creative control from recent research results and relate them to control mechanisms. By this we hope to further a more objective judging of the ability of a technique to support creativity and a detailed comparison of methods and give an first example for that in the following review of the state of the art.

Various aspects of our discussion guides still leave room for interpretation. Knowledge from other disciplines, for example in regard to the perception of visual features, can contribute with valuable insights. We hope that our work inspires such research towards a quantifiable analysis of creative control.

